---
title: "Gut microbiom Chronic Kidney Disease Incident"
author: "Joonatan Palmu, Li-Fang Yeo"
date: "17th Sept 2024"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE) 
```
```{r}
library(magrittr)
devtools::load_all()
```

```{r}
#Arguments
#makes it easier to run scripts (.sh) applicable when using emacs
args <- list(
  optparse::make_option("--east", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
  optparse::make_option("--prevalence", type = "numeric", default = 1/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()

args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
Formatting
```{r}
#Formatting how outputs are saved
myggsave <- myggsavefactory()

#Formatting image outputs
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(
      panel.border = ggplot2::element_rect(colour = "black", fill = NA),
      axis.text = ggplot2::element_text(colour = "black", size = 10),
      aspect.ratio = 1) } %>%
  ggplot2::theme_set()
```
Data definition
```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB = "Diabetes",
             SYSTM = "Systolic blood pressure",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT = "Heart failure",
             PREVAL_CKD = "Prevalent CKD",
             INCIDENT_CKD = "Incident CKD",
             CKD_AGEDIFF = "CKD Agediff",
             CURR_SMOKE = "Smoking",
             PREVAL_AUTOIMMUN = "Autoimmune disease",
             KREA_ENTS = "Creatinine",
             GFR = "Glomerulal filtration rate",
             UAC = "Urine Albumin-Creatinine Ratio",
             EAST = "Eastern Finland",
             shannon = "Shannon diversity",
             total_reads = "Total reads") 
```
Read in data, calculate alpha, tidy data for selected variables
```{r}
tse <- readRDS("../data/tse_mgs-20241118_104759.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::estimateDiversity(assay.type = "counts", index = "shannon", name = "shannon") %>%
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(GFR = 0.993^round(BL_AGE) *
               dplyr::case_when(MEN == 0 & KREA_ENTS <= 62 ~ 144*(KREA_ENTS/61.9)^-0.329,
                                MEN == 0 & KREA_ENTS >  62 ~ 144*(KREA_ENTS/61.9)^-1.209,
                                MEN == 1 & KREA_ENTS <= 80 ~ 141*(KREA_ENTS/79.6)^-0.411,
                                MEN == 1 & KREA_ENTS >  80 ~ 141*(KREA_ENTS/79.6)^-1.209)) %>%
  tse_mutate(UAC = U_ALB/U_KREA) %>% 
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>% 
  tse_filter(GFR >= 60, PREVAL_CKD == 0, UAC <=3 | is.na(UAC)) %>%
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars) %difference% "UAC"), not_na)) %>%
  tse_filter(total_reads > 50000) %>%
  tse_select(names(vars))
```
Table one / characteristics of data
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::mutate(INCIDENT_CKD = factor(ifelse(INCIDENT_CKD == 1, "Incident CKD" , "No CKD"))) %>%
  mytableone(vars,fo =  ~ .| INCIDENT_CKD )
```
Function use to define coxph model with all covariates
```{r}
coxph_full_partial <-  purrr::partial(survival::coxph,
                                      formula = my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~
                                      term + BL_AGE + MEN + BMI +
                                      PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                      PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                      PREVAL_AUTOIMMUN.col_from_endpoints + KREA_ENTS,
                                      ties = "breslow")

```
Function for cox model with minimum covariates (age + sex)
```{r}
coxph_minimum_partial <-  purrr::partial(survival::coxph,
                                         formula = my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~
                                           term + BL_AGE + MEN,
                                         ties = "breslow")
```
Function formating for missing variables
```{r}
cox_model_for_taxon <- function(df, term, fun, rank = NULL) {
  stopifnot(!missing(df), !missing(term), !missing(fun))
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    fun(data = .)
}
```
# ALPHA DIVERSITY
Using alpha diversity to predict incidence of CKD, adjusting only for sex and age
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>% 
  cox_model_for_taxon(term = "shannon", fun = coxph_minimum_partial) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Alpha diversity")
```
Using alpha diversity to predict incidence of CKD, adjusting for all covariates 
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  cox_model_for_taxon(term = "shannon", fun = coxph_full_partial) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  #DT::datatable(caption = "Alpha diversity")
  write.csv("cox-CKD.csv")
```

Common Taxa
```{r}
#Agglomerate prevalent features
subset_features_partial <-  purrr::partial(mia::subsetByPrevalentFeatures,
                                           x = tse,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```

```{r}
taxa_subsets <- c("Species") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```

```{r}
taxa_subsets %>%
  purrr::map_df(~mia::getPrevalence(.x, detection = args$detection, as_relative = TRUE), .id = "rank") %>%
  tidyr::gather(Taxa, Abundance, -rank) %>%
  dplyr::arrange(Abundance) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Gut microbial relative abundances at 0.1% detection threshold, in 1% of all samples")
```
Melt TSE to data frame
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltAssay(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
Cox's model
```{r}
#Check available taxa
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```

# Cox's model
Minimum (adjusted for age and sex)
```{r, message=FALSE}
taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~cox_model_for_taxon(df = dfs,
                                                                       term = .y,
                                                                       rank = .x,
                                                                       fun = coxph_minimum_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (minimum)")

```
Full (all covariates)
```{r, message= FALSE}
taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~cox_model_for_taxon(df = dfs,
                                                                       term = .y,
                                                                       rank = .x,
                                                                       fun = coxph_full_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (full)")
 
```
# Cox model by Genus
Common Taxa
```{r}
#Agglomerate prevalent features
subset_features_partial <-  purrr::partial(mia::subsetByPrevalentFeatures,
                                           x = tse,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```

```{r}
taxa_subsets <- c("Genus") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```

```{r}
taxa_subsets %>%
  purrr::map_df(~mia::getPrevalence(.x, detection = args$detection, as_relative = TRUE), .id = "rank") %>%
  tidyr::gather(Taxa, Abundance, -rank) %>%
  dplyr::arrange(Abundance) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Gut microbial relative abundances at 0.1% detection threshold, in 1% of all samples")
```
Melt TSE to data frame
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltAssay(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
Cox's model
```{r}
#Check available taxa
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```

# Cox's model
Minimum (adjusted for age and sex)
```{r, message=FALSE}
taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~cox_model_for_taxon(df = dfs,
                                                                       term = .y,
                                                                       rank = .x,
                                                                       fun = coxph_minimum_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (minimum)")
```
Full (all covariates)
```{r, message= FALSE}
taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~cox_model_for_taxon(df = dfs,
                                                                       term = .y,
                                                                       rank = .x,
                                                                       fun = coxph_full_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (full)")
```
# Beta diversity
Calculate RDA using Bray-curtis at the species level with only age and sex covariates
```{r}
#beta_ckd_full <- tse %>%  mia::agglomerateByRank(rank = "Species") %>% 
  #  mia::transformAssay(method = "relabundance") %>%
  #  mia::runRDA(assay.type = "relabundance",
   #             formula = assay ~ BL_AGE + MEN + BMI +
    #                                  PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
     #                                 PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
      #                                PREVAL_AUTOIMMUN.col_from_endpoints + KREA_ENTS + INCIDENT_CKD,
       #         distance = "bray",
        #        na.action = na.exclude)
```
```{r}
#rda_info <- attr(SingleCellExperiment::reducedDim(beta_ckd_full, "RDA"), "significance")

```
```{r}
#rda_info$permanova %>%
  #knitr::kable() #%>%
  #write.csv("./permanova/permanova-ckd.csv")
#rda_info$homogeneity %>%
  #knitr::kable() #%>%
  #write.csv("./permanova/homogeneity-ckd.csv")

```
