---
title: "CKD-prevalenceTaxa"
author: "Li-Fang Yeo"
date: "2024-06-18"
output: html_document
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE) 
```
```{r}
library(magrittr)
```
```{r}
devtools::load_all()
```
Command line arguments
```{r}
args <- list(
  optparse::make_option("--east", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--west", action="store_true", default=FALSE, help="Exclude Eastern Finland subpopulation [default \"%default\"]"),
  optparse::make_option("--low", action="store_true", default=FALSE, help="Exclude low GFR [default \"%default\"]"),
  optparse::make_option("--high", action="store_true", default=FALSE, help="Exclude high GFR [default \"%default\"]"),
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
optparse::make_option("--prevalence", type = "numeric", default = 1/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()
```
```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```
Formatting options
```{r}
mysave <- mysavefactory()
```
```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
  ggplot2::theme(
             panel.border = ggplot2::element_rect(colour = "black", fill = NA),
             axis.text = ggplot2::element_text(colour = "black", size = 10),
             aspect.ratio = 1) } %>%
  ggplot2::theme_set()

```
Data definition
```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB.col_from_endpoints = "Diabetes",
             SYSTM = "Systolic blood pressure",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT.col_from_endpoints = "Heart failure",
             PREVAL_CKD = "Prevalent CKD",
             INCIDENT_CKD = "Incident CKD",
             CKD_AGEDIFF = "CKD Agediff",
             CURR_SMOKE = "Smoking",
             PREVAL_AUTOIMMUN.col_from_endpoints  = "Autoimmune disease",
             KREA_ENTS = "Creatinine",
             GFR = "Glomerulal filtration rate",
             UAC = "Urine Albumin-Creatinine Ratio",
             EAST = "Eastern Finland",
             shannon = "Shannon diversity")
```
Read data in, calculate GFR, UAC and remove NAs
```{r}
tse <- readRDS("data/tse_gg2_MGS_FR02.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::addAlpha(assay.type = "counts", index = "shannon", name = "shannon") %>%
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(GFR = 0.993^round(BL_AGE) *
               dplyr::case_when(MEN == 0 & KREA_ENTS <= 62 ~ 144*(KREA_ENTS/61.9)^-0.329,
                                MEN == 0 & KREA_ENTS >  62 ~ 144*(KREA_ENTS/61.9)^-1.209,
                                MEN == 1 & KREA_ENTS <= 80 ~ 141*(KREA_ENTS/79.6)^-0.411,
                                MEN == 1 & KREA_ENTS >  80 ~ 141*(KREA_ENTS/79.6)^-1.209)) %>%
  tse_mutate(UAC = U_ALB/U_KREA) %>% 
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>%
  { if (args$low) tse_filter(., GFR >= 60) else . } %>%
  { if (args$high) tse_filter(., GFR < 60) else . } %>% 
  { if (args$east) tse_filter(., EAST == 0) else . } %>% 
  { if (args$west) tse_filter(., EAST == 1) else . } %>% 
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars) %difference% "UAC"), not_na)) %>%
  tse_select(names(vars))
```
# Characteristic
```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  mytableone(vars)
```
GFR model + covariates
```{r}
lm_gfr_partial <-  purrr::partial(lm,
                                  formula = GFR ~
                                    term + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints)
```
```{r}
lm_gfr_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_gfr_partial(data = .)
}
```
log(UAC) model + covariates
```{r}
lm_uac_partial <-  purrr::partial(lm,
                                  formula = log(UAC) ~
                                    term + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints)
```
```{r}
lm_uac_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_uac_partial(data = .)
}
```
Creatine model + covariates
```{r}
lm_crea_partial <-  purrr::partial(lm,
                                  formula = KREA_ENTS ~
                                    term + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints)
```
```{r}
lm_crea_model_for_taxon <- function(df, term, rank = NULL) {
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    lm_crea_partial(data = .)
}
```
# Taxa level analyses at ALL levels
#Common taxa
Agglomerate prevalent features
```{r}
subset_features_partial <-  purrr::partial(mia::subsetByPrevalentFeatures,
                                           x = tse,
                                           detection = args$detection,
                                           prevalence = args$prevalence,
                                           as_relative = TRUE) %>%
  purrr::possibly()
```
Subset to only Family
```{r}
taxa_subsets <- c("Genus") %>%
  rlang::set_names() %>%
  purrr::map(subset_features_partial)
```
Melt tse to data frame
```{r}
melt_tse <- function(x, method = "clr") {
  mia::transformAssay(x, method = method, pseudocount = 1) %>% 
    mia::meltAssay(add_row_data = TRUE, assay_name = method) %>%
    dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
    dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
    dplyr::select(SampleID, FeatureID, clr) %>%
    tidyr::spread(FeatureID, clr) %>% 
    dplyr::full_join(tse_meta(x), by = dplyr::join_by(SampleID == rownames))
}
```
```{r}
dfs <- taxa_subsets %>%
  purrr::map(melt_tse, .progress = TRUE)
```
# Taxa level analyses
generate a list of taxa, add the word GUT in front of the taxa
```{r}
taxa_rank_list <- function(x, y) {
  list_names <- colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names()
}
```
```{r}
taxa_rank_list <- dfs %>%
  purrr::imap(taxa_rank_list, .progress = TRUE)
```
## GFR
From the taxa list, add rank, and generate linear model for each taxa, adjust for FDR using PC method and BH
```{r, message=FALSE}
df_lm_gfr_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_gfr_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  #dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```
Results
```{r}
df_lm_gfr_results %>%
  dplyr::arrange(p.value) %>% 
  dplyr::filter(qval_fdr < 0.05) %>%
  DT::datatable(caption = "Linear model for GFR")
```
U-AlbKrea 
(Urine albumine creatine ratio)
```{r, message= FALSE}
df_lm_uac_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_uac_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  #dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```
```{r}
df_lm_uac_results %>%
  dplyr::arrange(p.value) %>% 
  dplyr::filter(qval_fdr < 0.05) %>%
  DT::datatable(caption = "Linear model for UAC")
```
Crea
 (serum creatine levels)
```{r,message=FALSE}
df_lm_crea_results <- taxa_rank_list %>%
  purrr::map_df(~tibble::tibble(taxa = .x), .id = "rank") %>% 
  tidyr::gather(rank, taxa) %>%
  dplyr::mutate(results = purrr::map2(rank, taxa, ~lm_crea_model_for_taxon(df = dfs, term = .y, rank = .x))) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  # dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```
```{r}
df_lm_crea_results %>%
  dplyr::arrange(p.value) %>% 
  dplyr::filter(qval_fdr < 0.05) %>%
  DT::datatable(caption = "Linear model for Serum Creatinine")
```
