---
title: "Functional Incident CKD"
author: "LF Yeo"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message=FALSE) 
```
```{r}
library(magrittr)
library(RNOmni)
library(dplyr)
library(readr)
library(mia)
devtools::load_all()
```
Importing files
```{r}
#Functional pathways from metaCyc
joined_pathabundance <- read_delim("joined_pathabundance.tsv", 
                                   delim = "\t", escape_double = FALSE, trim_ws = TRUE)
```
```{r}
#Defining covariates for the model
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB.col_from_endpoints = "Diabetes",
             SYSTM = "Systolic blood pressure",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT.col_from_endpoints = "Heart failure",
             PREVAL_CKD = "Prevalent CKD",
             INCIDENT_CKD = "Incident CKD",
             CKD_AGEDIFF = "CKD Agediff",
             CURR_SMOKE = "Smoking",
             PREVAL_AUTOIMMUN.col_from_endpoints = "Autoimmune disease",
             KREA_ENTS = "Creatinine",
             GFR = "Glomerulal filtration rate",
             UAC = "Urine Albumin-Creatinine Ratio",
             EAST = "Eastern Finland") 

#Importing metadata and cleaning up
tse <- readRDS("data/tse_gg2_MGS_FR02.rds") %>%
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(GFR = 0.993^round(BL_AGE) *
               dplyr::case_when(MEN == 0 & KREA_ENTS <= 62 ~ 144*(KREA_ENTS/61.9)^-0.329,
                                MEN == 0 & KREA_ENTS >  62 ~ 144*(KREA_ENTS/61.9)^-1.209,
                                MEN == 1 & KREA_ENTS <= 80 ~ 141*(KREA_ENTS/79.6)^-0.411,
                                MEN == 1 & KREA_ENTS >  80 ~ 141*(KREA_ENTS/79.6)^-1.209)) %>%
  tse_mutate(UAC = U_ALB/U_KREA) %>% 
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>% 
  tse_filter(GFR >= 60, PREVAL_CKD == 0, UAC <=3 | is.na(UAC)) %>%
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars) %difference% "UAC"), not_na)) %>%
  tse_select(names(vars))
```
Data wrangling and cleaning up
```{r}
#Matching sampleID from functional data to metadata
#rename 1st column name
a <- as.data.frame(colData(tse))%>%
  tibble::rownames_to_column("Barcode")
# Get the column names (excluding the first one, which is probably the path/feature column)
sample_names <- colnames(joined_pathabundance)[-1]

# Clean sample names by removing after .R1
cleaned_sample_names <- gsub("\\.R1.*","", sample_names, perl = TRUE)

# Update the column names with the cleaned sample names
colnames(joined_pathabundance)[-1] <- cleaned_sample_names

# Find the intersection of sample names between the cleaned joined_path_abundance sample names and FINRISK sample data
common_samples <- intersect(a$Barcode, cleaned_sample_names)

# Subset the joined_pathabundance file to include only the common samples
joined_pathabundance<- joined_pathabundance %>%
  select(all_of(c("Pathway", common_samples)))

rm(tse)
```
Filtering any pathways associated w/ species
```{r}
#Analysis focused on finding general pathways
shit<- joined_pathabundance %>% 
  filter(rowSums(across(where(is.numeric)))!=0) %>% 
  filter(!grepl('.s__', Pathway))
```
Calculating coefficient of variance
```{r}
#calculate standard coefficient variation by row-wise feature
cv <- shit %>%
  rowwise() %>%
  mutate(
    mean_val = mean(c_across(2:6567)),
    sd_val = sd(c_across(2:6567)),
    cv = (sd_val / mean_val) * 100) %>%
  ungroup()

hist(cv$cv)
```
Transpose dataframe
```{r}
shit.t <- as.data.frame(t(shit))
#take first row as colname
colnames(shit.t) <- shit.t[1,]
#remove extra row
shit.t <- shit.t[-1, ] 
#change from character to numeric
shit.t <- shit.t %>% mutate_if(is.character, as.numeric) %>%
  tibble::rownames_to_column("Barcode") 
```
Reduce to dichotomous variables (absence / presence)
```{r}
poo_dicho <- shit.t %>%
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, 0)))
```
Data wrangling to join functional data to metadata
```{r}
df <- dplyr::inner_join(poo_dicho, a, by = "Barcode")
```
Cox model for testing pathways associated with Incident CKD (outcome variable)
```{r}
##COX model
coxph_full_partial <-  purrr::partial(survival::coxph,
                                      formula = my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~
                                        term + BL_AGE + MEN + BMI +
                                        PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                        PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                        PREVAL_AUTOIMMUN.col_from_endpoints + KREA_ENTS,
                                      ties = "breslow") %>%
                      purrr::possibly()

#Take the column names from Pathway in shit dataframe as a list and mutate
try <- shit %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try%>% 
  dplyr::filter(!is.na(results)) %>%
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")

try%>% 
  dplyr::filter(is.na(results))
```
Tranform to inverse_rank
```{r}
# Define the transformation function
transform_column <- function(x) {
  qnorm((rank(x, na.last = "keep") - 0.5) / sum(!is.na(x)))
}

# Apply the transformation to all columns
poo_rank <- shit.t
poo_rank[, -1] <- lapply(shit.t[, -1], transform_column)
```
Data wrangling to join functional data to metadata
```{r}
df <- dplyr::inner_join(poo_rank, a, by = "Barcode")
```
```{r}
##Cox for inverse rank
#Take the column names from Pathway in shit dataframe as a list and mutate
try <- shit %>% dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE))

try%>% dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")
```
Remove any samples with zero
```{r}
#Take the column names from Pathway in shit dataframe as a list and mutate
df <- shit.t %>% 
  mutate_if(is.numeric, ~ifelse(abs(.) < 10^-10, NA, .)) %>%
  dplyr::inner_join(a, by = "Barcode")

try <- shit %>% 
  dplyr::select(Pathway) %>% 
  dplyr::mutate(results = purrr::map(Pathway, ~coxph_full_partial(data = df %>% dplyr::rename(term = .x)), .progress = TRUE, na.rm = TRUE))

try%>% dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(term == "term") %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH")) %>%
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 3)) %>%
  DT::datatable(caption = "Cox model for MetaCyc pathways")
```


