---
title: "Prevalent nephro-metagenomics in FINRISK 2002"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  rmdformats::html_clean:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    code_folding: show
---

# Setup

```{css, echo = FALSE}
.main-container {
  max-width:1920px !important;
}
```

```{r libraries}
library(magrittr)
```

```{r load functions}
devtools::load_all()
```

## Command line arguments

```{r}
opts <- list(
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
  optparse::make_option("--prevalence", type = "numeric", default = 1/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()
```

```{r}
opts %>% tibble::enframe(name = "Option", value = "Argument") %>% knitr::kable()
```

## Formatting options

```{r options, echo = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, results='asis', cache=FALSE, warning=FALSE, error = TRUE)
```

```{r define ggplot settings}
mysave <- mysavefactory()
```

```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
  ggplot2::theme(
             panel.border = ggplot2::element_rect(colour = "black", fill = NA),
             axis.text = ggplot2::element_text(colour = "black", size = 10),
             aspect.ratio = 1) } %>%
  ggplot2::theme_set()
```
    
# Data definitions

```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB.col_from_endpoints = "Diabetes",
             SYSTM = "Systolic blood pressure",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT.col_from_endpoints = "Heart failure",
             PREVAL_CKF = "Prevalent CKD",
             INCIDENT_CKF = "Incident CKD",
             CKF_AGEDIFF = "CKD Agediff",
             CURR_SMOKE = "Smoking",
             PREVAL_AUTOIMMUN.col_from_endpoints  = "Autoimmune disease",
             KREA_ENTS = "Creatinine",
             GFR = "Glomerulal filtration rate",
             UAC = "Urine Albumin-Creatinine Ratio",
             EAST = "Eastern Finland",
             CKD_STAGE = "Stage of chronic kidney disease",
             shannon = "Shannon diversity") 
```

```{r}
tse <- readRDS("data/tse_gg2_MGS_FR02.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::estimateDiversity(assay.type = "counts", index = "shannon", name = "shannon") %>%
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(GFR = 0.993^round(BL_AGE) *
               dplyr::case_when(MEN == 0 & KREA_ENTS <= 62 ~ 144*(KREA_ENTS/61.9)^-0.329,
                                MEN == 0 & KREA_ENTS >  62 ~ 144*(KREA_ENTS/61.9)^-1.209,
                                MEN == 1 & KREA_ENTS <= 80 ~ 141*(KREA_ENTS/79.6)^-0.411,
                                MEN == 1 & KREA_ENTS >  80 ~ 141*(KREA_ENTS/79.6)^-1.209)) %>%
  tse_mutate(CKD_STAGE = factor(dplyr::case_when(GFR >= 90 ~ "Stage 1",
                                                 dplyr::between(GFR, 60,  90) ~ "Stage 2",
                                                 dplyr::between(GFR, 30,  60) ~ "Stage 3",
                                                 dplyr::between(GFR, 15,  30) ~ "Stage 4",
                                                 GFR < 15 ~ "Stage 5"))) %>% 
  tse_mutate(UAC = U_ALB/U_KREA) %>% 
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>%
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars) %difference% "UAC"), not_na)) %>%
  tse_select(names(vars))
```

# Characteristics

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::mutate(EAST = ifelse(EAST == 1, "East", "West")) %>% 
  mytableone(vars, fo = ~. | EAST)
```

## Distributions

```{r}
{ tse %>%
    tse_meta(rownames = FALSE) %>%
    ggplot2::ggplot(ggplot2::aes(x = CKD_STAGE, fill = EAST)) +
    ggplot2::geom_bar(stat = "count") +
    ggplot2::geom_text(stat='count',
                       ggplot2::aes(label = ..count..),
                       color = "black",
                       position = ggplot2::position_stack(vjust = 0.5)) +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::scale_fill_grey(start = 0.3, end = 0.5) +
    ggplot2::scale_y_continuous(name = NULL, expand = c(0, 0)) +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) } %>%
  mysave(name = "nephrobiota_gfr_distribution", width = 8, height = 7)
```

# Model function definitions

```{r}
bin_krea_partial <-  purrr::partial(glm,
                                    formula = CKD_STAGE ~
                                    term + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints,
                                    family = binomial)
```


```{r}
lm_krea_partial <-  purrr::partial(lm,
                                  formula = KREA_ENTS ~
                                    term + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints)
```

```{r}
lm_uac_partial <-  purrr::partial(lm,
                                  formula = log(UAC) ~
                                    term + BL_AGE + MEN + BMI +
                                      PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                      PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                      PREVAL_AUTOIMMUN.col_from_endpoints)
```

```{r}
cox_model_for_taxon <- function(df, term, fun, rank = NULL) {
  stopifnot(!missing(df), !missing(term), !missing(fun))
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    fun(data = .)
}
```

# Alpha diversity

## Univariate

### Crea

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  lm(KREA_ENTS ~ shannon, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>%
  knitr::kable(caption = "Alpha diversity and Crea")
```

### UAC

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::filter(!is.na(UAC)) %>% 
  lm(log(UAC) ~ shannon, data = .) %>%
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>%
  knitr::kable(caption = "Alpha diversity and UAC")
```

## Multivariate

### Crea

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  lm(KREA_ENTS ~ shannon + BL_AGE + MEN + BMI +
       PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
       PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
       PREVAL_AUTOIMMUN.col_from_endpoints, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  knitr::kable(caption = "Alpha diversity and Crea")
```

### UAC

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::filter(!is.na(UAC)) %>% 
  lm(UAC ~ shannon + BL_AGE + MEN + BMI +
       PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
       PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
       PREVAL_AUTOIMMUN.col_from_endpoints, data = .) %>% 
  broom::tidy(conf.int = TRUE) %>%
  round_numeric_columns() %>% 
  knitr::kable(caption = "Alpha diversity and UAC")
```

# Approximate microbial data dimensions

```{r}
pca <- tse %>%
  tse_filter(EAST == 1) %>%
  mia::subsetByPrevalentFeatures(detection = opts$detection, prevalence = opts$prevalence, as_relative = TRUE) %>% 
  SummarizedExperiment::assay("relabundance") %>%
  t %>%
  tibble::as_tibble() %>%
  prcomp()
```

```{r}
n_independent_axes <- broom::tidy(pca, matrix = "eigenvalues") %>%
  dplyr::filter(cumulative < 0.9) %>%
  nrow %>%
  add(1)
```

N:o axes `r n_independent_axes`.

# Common taxa

## East

```{r}
tse_species_east <- mia::subsetByPrevalentFeatures(x = tse %>% tse_filter(EAST == 1),
                                              rank = "Species",
                                              detection = opts$detection,
                                              prevalence = opts$prevalence,
                                              as_relative = TRUE)
```

```{r}
df_east <- tse_species_east %>%
  mia::transformAssay(method = "clr", pseudocount = 1) %>% 
  mia::meltAssay(add_row_data = TRUE, assay_name = "clr") %>%
  dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
  dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
  dplyr::select(SampleID, FeatureID, clr) %>%
  tidyr::spread(FeatureID, clr) %>% 
  dplyr::full_join(tse_meta(tse_species_east), by = dplyr::join_by(SampleID == rownames))
```

## West

```{r}
tse_species_west <- mia::subsetByPrevalentFeatures(x = tse %>% tse_filter(EAST == 0),
                                                   rank = "Species",
                                                   detection = 0,
                                                   prevalence = 0,
                                                   as_relative = TRUE)
```

```{r}
df_west <- tse_species_west[rownames(tse_species_east), ] %>%
  mia::transformAssay(method = "clr", pseudocount = 1) %>% 
  mia::meltAssay(add_row_data = TRUE, assay_name = "clr") %>%
  dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
  dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
  dplyr::select(SampleID, FeatureID, clr) %>%
  tidyr::spread(FeatureID, clr) %>% 
  dplyr::full_join(tse_meta(tse_species_west), by = dplyr::join_by(SampleID == rownames))
```

# Taxa level analyses

## Check available taxa

```{r}
taxa_rank_list <- function(x, y) {
  colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names() %>%
    tibble::tibble(taxa = .)
}
```

## Linear models

### Crea

```{r}
df_lm_krea_east <- taxa_rank_list(df_east) %>% 
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_east,
                                                                term = .x,
                                                                fun = lm_krea_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
df_lm_krea_east %>% 
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (minimum)")
```

### U-AlbKrea

```{r}
df_lm_uac_east <- taxa_rank_list(df_east) %>% 
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_east %>% dplyr::filter(!is.na(UAC)),
                                                                term = .x,
                                                                fun = lm_uac_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
df_lm_uac_east %>% 
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (minimum)")
```

## Validation

### Crea

```{r}
df_lm_krea_west <- df_lm_krea_east %>%
  dplyr::filter(qval_my < 0.05) %>%
  dplyr::select(taxa) %>%
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_west,
                                                                term = .x,
                                                                fun = lm_krea_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  { if ("term" %in% colnames(.)) . else tibble::add_column(., term = NA_character_, p.value = NA_real_) } %>% 
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = dplyr::n())) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
dplyr::inner_join(df_lm_krea_east, df_lm_krea_west, by = dplyr::join_by(taxa, term), suffix = c("_east", "_west")) %>%
  dplyr::select(contains("taxa"), contains("estimate"), contains("qval")) %>% 
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>%
  dplyr::arrange(qval_my_west) %>% 
  DT::datatable()
```

### UAC

```{r}
df_lm_uac_west <- df_lm_uac_east %>%
  dplyr::filter(qval_my < 0.05) %>%
  dplyr::select(taxa) %>%
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_west %>% dplyr::filter(!is.na(UAC)),
                                                                term = .x,
                                                                fun = lm_uac_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, conf.int = TRUE))) %>%
  tidyr::unnest(results) %>%
  { if ("term" %in% colnames(.)) . else tibble::add_column(., term = NA_character_, p.value = NA_real_) } %>% 
  dplyr::filter(stringr::str_detect(term, "term")) %>% 
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = dplyr::n())) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
dplyr::inner_join(df_lm_uac_east, df_lm_uac_west, by = dplyr::join_by(taxa, term), suffix = c("_east", "_west")) %>%
  dplyr::select(contains("taxa"), contains("estimate"), contains("qval")) %>% 
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable()
```

# Risk score

```{r}
taxa_for_glmnet <- df_lm_krea_east %>%
  dplyr::filter(p.value < 0.05) %>%
  dplyr::pull(taxa)
```

```{r}
glmnet_x <- glue::glue("KREA_ENTS ~ 0 + BL_AGE + MEN + BMI +
                                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                                    PREVAL_AUTOIMMUN.col_from_endpoints +
                                    {stringr::str_c(taxa_for_glmnet, collapse = ' + ')}") %>% 
  { model.matrix(as.formula(.), df_east) }
glmnet_y <- df_east %>% dplyr::pull(KREA_ENTS)
```

```{r}
set.seed(20240305)
cox_model <- glmnet::cv.glmnet(glmnet_x,
                               glmnet_y,
                               maxit=10**5,
                               alpha = 1)
```

```{r}
plot(cox_model)
```


## Implement risk score

```{r}
df_glmnet <- coef(cox_model, s = "lambda.min") %>%
  as.data.frame.matrix() %>% 
  tibble::as_tibble(rownames = "term") %>%
  dplyr::filter(stringr::str_detect(term, "GUT_")) %>%
  dplyr::mutate(value = round(s1, 2)) %>% 
  dplyr::filter(abs(value) > 0) %>%
  dplyr::select(term, value)
```

### Formula coefficients

```{r}
{ df_glmnet %>%
    dplyr::mutate(term = stringr::str_replace_all(term, c("GUT_" = "", "_" = " "))) %>% 
    ggplot2::ggplot(ggplot2::aes(x = term, y = value)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::scale_y_continuous(name = "Weight to risk score", expand = c(0, 0)) +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) +
    ggplot2::coord_flip() +
    ggplot2::theme(legend.position = "none",
                   panel.grid.major = ggplot2::element_blank(),
                   panel.grid.minor = ggplot2::element_blank(),
                   plot.margin = ggplot2::margin(1, 5, 1, 1, "mm")) } %>% 
  mysave(name = "nephrobiota_risk_features", width = 10, height = 5)
```

### Risk formula

```{r}
fo <- df_glmnet %>%
  dplyr::mutate(fo = glue::glue("({value}*{term})")) %>%
  dplyr::pull(fo) %>%
  stringr::str_c(collapse = "+")
```

```{r, results = "asis"}
fo
```


### Models

```{r}
dfs_risk <- list()
```

#### Discovery

```{r}
( dfs_risk$train <- df_east %>% 
    dplyr::mutate(risk := rlang::eval_tidy(rlang::parse_expr(fo))) %>% 
    cox_model_for_taxon(term = "risk", fun = lm_krea_partial) %>% 
    broom::tidy(conf.int = TRUE) ) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Risk score full model")
```

#### Validation

```{r}
( dfs_risk$test <- df_west %>% 
    dplyr::mutate(risk := rlang::eval_tidy(rlang::parse_expr(fo))) %>%
    cox_model_for_taxon(term = "risk", fun = lm_krea_partial) %>% 
    broom::tidy(conf.int = TRUE) ) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Risk score full model")
```

### Combine results

```{r}
df_risk_plot <- vars %>%
    tibble::enframe(value = "label") %>%
    tidyr::unnest(label) %>%
    tibble::add_row(name = "term", label = "Gut microbial risk score") %>%
    dplyr::right_join(dplyr::bind_rows(dfs_risk, .id = "model") %>%
                      dplyr::mutate(name = stringr::str_replace_all(term,
                                                                    stringr::regex("[0-9]$"), "")),
                      by = dplyr::join_by(name)) %>%
    dplyr::filter(term != "(Intercept)")  
```


```{r}
{ ggplot2::ggplot(df_risk_plot,
                    ggplot2::aes(x = label, y = estimate, color = model)) +
    ggplot2::geom_errorbar(ggplot2::aes(ymin = conf.low, ymax = conf.high),
                           position = ggplot2::position_dodge(width = .5),
                           width=0.4) +
    ggplot2::geom_hline(yintercept = 0) +
    ggplot2::geom_point(position = ggplot2::position_dodge(width = .5)) +
    ggplot2::scale_color_grey(start = 0, end = 0.4) +
    ggplot2::scale_y_continuous(name = NULL, expand = c(0, 0), limits = ~max(abs(.)+1) %>% c(-., .)) +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) +
    ggplot2::coord_flip() } %>%
  mysave("nephrobiota_risk_score", width = 10, height = 4)  
```
