---
title: "Nephro-metagenomics in FINRISK 2002"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  rmdformats::html_clean:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
---

# Setup

```{css, echo = FALSE}
.main-container {
  max-width:1920px !important;
}
```

```{r libraries}
library(magrittr)
```

```{r load functions}
devtools::load_all()
```

## Command line arguments

```{r}
args <- list(
  optparse::make_option("--detection", type = "numeric", default = 0.1/100, help = "Detection limit [default %default]"),
  optparse::make_option("--prevalence", type = "numeric", default = 1/100, help = "Prevalence limit [default %default]")) %>% 
  optparse::OptionParser(option_list = .) %>%
  optparse::parse_args()
```

```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% DT::datatable()
```

## Formatting options

```{r options, echo = FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, results='asis', cache=FALSE, warning=FALSE, error = TRUE)
```

    
```{r define ggplot settings}
myggsave <- myggsavefactory()
```

```{r}
{ ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
    ggplot2::theme(
               panel.border = ggplot2::element_rect(colour = "black", fill = NA),
               axis.text = ggplot2::element_text(colour = "black", size = 10),
               aspect.ratio = 1) } %>%
  ggplot2::theme_set()
``` 
    
# Data definitions

```{r}
vars <- list(BL_AGE = "Age",
             MEN = "Men",
             BMI = "BMI",
             PREVAL_DIAB.col_from_endpoints = "Diabetes",
             SYSTM = "Systolic blood pressure",
             BP_TREAT = "Antihypertensive medication",
             PREVAL_HFAIL_STRICT.col_from_endpoints = "Heart failure",
             PREVAL_CKD = "Prevalent CKD",
             INCIDENT_CKD = "Incident CKD",
             CKD_AGEDIFF = "CKD Agediff",
             CURR_SMOKE = "Smoking",
             PREVAL_AUTOIMMUN.col_from_endpoints = "Autoimmune disease",
             KREA_ENTS = "Creatinine",
             GFR = "Glomerulal filtration rate",
             UAlbKrea = "Urine Albumin-Creatinine Ratio",
             EAST = "Eastern Finland",
             shannon = "Shannon diversity") 
```

```{r}
tse <- readRDS("data/tse_gg2_MGS_FR02.rds") %>%
  mia::transformAssay(assay.type = "counts", method = "relabundance") %>% 
  mia::estimateDiversity(assay.type = "counts", index = "shannon", name = "shannon") %>%
  tse_mutate(dplyr::across(c(MEN,
                             EAST,
                             BP_TREAT,
                             CURR_SMOKE,
                             dplyr::contains("INCIDENT"),
                             dplyr::contains("PREVAL")), as.factor)) %>% 
  tse_mutate(GFR = 0.993^round(BL_AGE) *
               dplyr::case_when(MEN == 0 & KREA_ENTS <= 62 ~ 144*(KREA_ENTS/61.9)^-0.329,
                                MEN == 0 & KREA_ENTS >  62 ~ 144*(KREA_ENTS/61.9)^-1.209,
                                MEN == 1 & KREA_ENTS <= 80 ~ 141*(KREA_ENTS/79.6)^-0.411,
                                MEN == 1 & KREA_ENTS >  80 ~ 141*(KREA_ENTS/79.6)^-1.209)) %>%
  tse_mutate(UAlbKrea = U_ALB/U_KREA) %>% 
  tse_filter(GRAVID %in% c(1, NA), BL_USE_RX_J01_1mo %in% c(0, NA)) %>% 
  tse_filter(GFR >= 60, PREVAL_CKD == 0, UAlbKrea <=3 | is.na(UAlbKrea)) %>%
  tse_filter(dplyr::if_all(dplyr::one_of(names(vars) %difference% "UAlbKrea"), not_na)) %>%
  tse_select(names(vars))
```

# Characteristics

```{r}
tse %>%
  tse_meta(rownames = FALSE) %>%
  dplyr::mutate(EAST = ifelse(EAST == 1, "East", "West")) %>% 
  mytableone(vars, fo = ~. | EAST)
```

# Model definitions

```{r}
coxph_full_partial <-  purrr::partial(survival::coxph,
                                      formula = my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~
                                        term + BL_AGE + MEN + BMI +
                                          PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                                          CURR_SMOKE,
                                      ties = "breslow")
```

```{r}
coxph_minimum_partial <-  purrr::partial(survival::coxph,
                                      formula = my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~
                                        term + BL_AGE + MEN,
                                      ties = "breslow")
```

```{r}
cox_model_for_taxon <- function(df, term, fun, rank = NULL) {
  stopifnot(!missing(df), !missing(term), !missing(fun))
  message(term)
  { if (is.null(rank)) df else df[[rank]] } %>%
    dplyr::rename(term := {{term}}) %>% 
    fun(data = .)
}
```


# Alpha diversity

## Univariate

```{r}
tse %>%
  tse_filter(EAST == 1) %>% 
  tse_meta(rownames = FALSE) %>% 
  cox_model_for_taxon(term = "shannon", fun = coxph_minimum_partial) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Alpha diversity")
```

## Multivariate

```{r}
tse %>%
  tse_filter(EAST == 1) %>% 
  tse_meta(rownames = FALSE) %>%
  cox_model_for_taxon(term = "shannon", fun = coxph_full_partial) %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Alpha diversity")
```

# Approximate microbial data dimensions

```{r}
pca <- tse %>%
  tse_filter(EAST == 1) %>%
  mia::subsetByPrevalentFeatures(detection = args$detection, prevalence = args$prevalence, as_relative = TRUE) %>% 
  SummarizedExperiment::assay("relabundance") %>%
  t %>%
  tibble::as_tibble() %>%
  prcomp()
```

```{r}
n_independent_axes <- broom::tidy(pca, matrix = "eigenvalues") %>%
  dplyr::filter(cumulative < 0.9) %>%
  nrow %>%
  add(1)
```

```{r, fig.cap="PCA for bacterial counts", out.width="50%", echo = FALSE}
{ broom::tidy(pca, matrix = "eigenvalues") %>%
    ggplot2::ggplot(ggplot2::aes(x = PC))+
    ggplot2::geom_bar(ggplot2::aes(y = cumulative),
             stat = "identity",
             fill = "steelblue",
             colour = "steelblue",
             width = 1) +
    ggplot2::geom_bar(ggplot2::aes(y = percent),
             stat = "identity",
             fill = "red",
             colour = "red",
             width = 1) +
    ggplot2::geom_hline(yintercept = 0.90, linetype = 2) +
    ggplot2::geom_vline(xintercept = n_independent_axes, linetype = 2) +
    ggplot2::scale_y_continuous(name = "Proportion of variance",
                       labels = scales::percent_format(accuracy = 1),
                       breaks = seq(0, 1, 0.5) %union% c(0.9),
                       expand = c(0, 0)) +
    ggplot2::scale_x_continuous(name = "PC axes",
                       breaks = seq(0, 1000, 50) %union% c(n_independent_axes),
                       expand = c(0, 0)) +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1))
  } %>%
  myggsave(name = "gut_pca_variance", width = 5, height = 5)
```

# Common taxa

## Agglomerate prevalent features

```{r}
tse_species <- mia::subsetByPrevalentFeatures(x = tse,
                                              rank = "Species",
                                              detection = args$detection,
                                              prevalence = args$prevalence,
                                              as_relative = TRUE)
```

```{r}
tse_species %>%
  mia::getPrevalence(detection = args$detection, as_relative = TRUE) %>%
  tibble::enframe(name = "Taxa", value = "Abundance") %>%
  dplyr::arrange(-Abundance) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Gut microbial abundances")
```

## Melt TSE to data frame

```{r}
df <- tse_species %>%
  mia::transformAssay(method = "clr", pseudocount = 1) %>% 
  mia::meltAssay(add_row_data = TRUE, assay_name = "clr") %>%
  dplyr::mutate(FeatureID = glue::glue("GUT_{FeatureID}")) %>%
  dplyr::mutate(FeatureID = stringr::str_replace_all(FeatureID, c(" " = "_", "-" = "_"))) %>% 
  dplyr::select(SampleID, FeatureID, clr) %>%
  tidyr::spread(FeatureID, clr) %>% 
  dplyr::full_join(tse_meta(tse_species), by = dplyr::join_by(SampleID == rownames))
```

```{r}
df_training <- df %>% dplyr::filter(EAST == 1)
df_test <- df %>% dplyr::filter(EAST == 0)
```

# Cox's models

## Check available taxa

```{r}
taxa_rank_list <- function(x, y) {
  colnames(x) %>%
    stringr::str_subset("GUT_") %>%
    rlang::set_names() %>%
    tibble::tibble(taxa = .)
}
```

## Cox's models

### Age and sex adjusted model

```{r}
df_cox_minimum_east <- taxa_rank_list(df_training) %>% 
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_training,
                                                                term = .x,
                                                                fun = coxph_minimum_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
df_cox_minimum_east %>% 
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (minimum)")
```

### Full adjusted model

```{r}
df_cox_full_east <- taxa_rank_list(df_training) %>% 
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_training,
                                                                term = .x,
                                                                fun = coxph_full_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
df_cox_full_east %>% 
  dplyr::arrange(p.value) %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Per taxa Cox's models (full)")
```

## Validation

### Age and sex adjusted model

```{r}
df_cox_minimum_west <- df_cox_minimum_east %>%
  dplyr::filter(qval_my < 0.05) %>%
  dplyr::select(taxa) %>%
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_test,
                                                                term = .x,
                                                                fun = coxph_minimum_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```

```{r}
dplyr::inner_join(df_cox_minimum_east, df_cox_minimum_west, by = dplyr::join_by(taxa, term), suffix = c("_east", "_west")) %>%
  dplyr::select(contains("taxa"), contains("estimate"), contains("p.value")) %>% 
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable()
```

### Full adjusted model

```{r}
df_cox_full_west <- df_cox_full_east %>%
  dplyr::filter(qval_my < 0.05) %>%
  dplyr::select(taxa) %>% 
  dplyr::mutate(results = purrr::map(taxa, ~cox_model_for_taxon(df = df_test,
                                                                term = .x,
                                                                fun = coxph_full_partial), .progress = TRUE)) %>% 
  dplyr::mutate(results = purrr::map(results, ~broom::tidy(.x, exponentiate = TRUE, conf.int = TRUE)))  %>%
  tidyr::unnest(results) %>%
  dplyr::filter(stringr::str_detect(term, "term")) %>%
  dplyr::mutate(qval_my = my_adjust_p(p.value, n = n_independent_axes)) %>%
  dplyr::mutate(qval_fdr = p.adjust(p.value, method = "BH"))
```


```{r}
dplyr::inner_join(df_cox_full_east, df_cox_full_west, by = dplyr::join_by(taxa, term), suffix = c("_east", "_west")) %>%
  dplyr::select(contains("taxa"), contains("estimate"), contains("p.value")) %>% 
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable()
```

# Risk score

```{r}
taxa_for_glmnet <- df_cox_minimum_east %>%
  dplyr::filter(p.value < 0.05) %>%
  dplyr::pull(taxa)
```

```{r}
glmnet_x <- glue::glue("my_surv(CKD_AGEDIFF, INCIDENT_CKD) ~ 0 +
                    BL_AGE + MEN + BMI +
                    PREVAL_DIAB.col_from_endpoints + SYSTM + BP_TREAT +
                    PREVAL_HFAIL_STRICT.col_from_endpoints + CURR_SMOKE +
                    PREVAL_AUTOIMMUN.col_from_endpoints + {stringr::str_c(taxa_for_glmnet, collapse = ' + ')}") %>% 
  { model.matrix(as.formula(.), df_training) }
glmnet_y <- df_training %$% my_surv(CKD_AGEDIFF, INCIDENT_CKD)
```

```{r}
set.seed(20240305)
cox_model <- glmnet::cv.glmnet(glmnet_x,
                               glmnet_y,
                               family = "cox",
                               maxit=10**5,
                               alpha = 1)
```

```{r}
plot(cox_model)
```


## Implement risk score

```{r}
df_glmnet <- coef(cox_model, s = "lambda.min") %>%
  as.data.frame.matrix() %>% 
  tibble::as_tibble(rownames = "term") %>%
  dplyr::filter(stringr::str_detect(term, "GUT_")) %>%
  dplyr::mutate(value = round(`1`, 2)) %>% 
  dplyr::filter(abs(value) > 0) %>%
  dplyr::select(term, value)
```

### Formula coefficients

```{r}
{ df_glmnet %>%
    dplyr::mutate(term = stringr::str_replace_all(term, c("GUT_" = "", "_" = " "))) %>% 
    ggplot2::ggplot(ggplot2::aes(x = term, y = value)) +
    ggplot2::geom_bar(stat = "identity") +
    ggplot2::scale_y_continuous(name = "Weight to risk score", expand = c(0, 0)) +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) +
    ggplot2::coord_flip() +
    ggplot2::theme(legend.position = "none",
                   panel.grid.major = ggplot2::element_blank(),
                   panel.grid.minor = ggplot2::element_blank(),
                   plot.margin = ggplot2::margin(1, 5, 1, 1, "mm")) } %>% 
  myggsave(name = "nephrobiota_risk_features", width = 10, height = 5)
```

### Risk formula

```{r}
fo <- df_glmnet %>%
  dplyr::mutate(fo = glue::glue("({value}*{term})")) %>%
  dplyr::pull(fo) %>%
  stringr::str_c(collapse = "+")
```

### Models

```{r}
dfs_risk <- list()
```

#### Discovery

```{r}
( dfs_risk$train <- df_training %>% 
    dplyr::mutate(risk := rlang::eval_tidy(rlang::parse_expr(fo))) %>% 
    cox_model_for_taxon(term = "risk", fun = coxph_full_partial) %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) ) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Risk score full model")
```

#### Validation

```{r}
( dfs_risk$test <- df_test %>% 
    dplyr::mutate(risk := rlang::eval_tidy(rlang::parse_expr(fo))) %>%
    cox_model_for_taxon(term = "risk", fun = coxph_full_partial) %>% 
    broom::tidy(exponentiate = TRUE, conf.int = TRUE) ) %>%
  round_numeric_columns() %>%
  dplyr::mutate_if(is.numeric, ~round(.x, digits = 4)) %>% 
  DT::datatable(caption = "Risk score full model")
```

### Combine results

```{r}
( df_risk_plot <- vars %>%
  tibble::enframe(value = "label") %>%
  tidyr::unnest(label) %>%
  tibble::add_row(name = "term", label = "Gut microbial risk score") %>%
  dplyr::right_join(dplyr::bind_rows(dfs_risk, .id = "model") %>%
                    dplyr::mutate(name = stringr::str_replace_all(term,
                                                                  stringr::regex("[0-9]$"), "")),
                    by = dplyr::join_by(name)) ) %>%
  DT::datatable()
```


```{r}
{ ggplot2::ggplot(df_risk_plot,
                    ggplot2::aes(x = label, y = estimate, color = model)) +
    ggplot2::geom_errorbar(ggplot2::aes(ymin = conf.low, ymax = conf.high),
                           position = ggplot2::position_dodge(width = .5),
                           width=0.4) +
    ggplot2::geom_hline(yintercept = 1) +
    ggplot2::geom_point(position = ggplot2::position_dodge(width = .5)) +
    ggplot2::scale_color_grey(start = 0, end = 0.4) +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::coord_flip() } %>%
  myggsave("risk_score", width = 10, height = 4)  
```
