---
title: "XGboost for the feature importance for each potentially causally linked metabolic measure"
author: "Joonatan Palmu"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  rmdformats::html_clean:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
---

# Init

```{css, echo = FALSE}
.page-inner, #content{
  max-width:1920px !important;
}
``` 


```{r options, include=FALSE}
knitr::opts_chunk$set(include = TRUE, echo = TRUE, message = FALSE, results='asis', cache=FALSE, warning=FALSE, collapse = TRUE)
```

# Library

```{r}
library(magrittr)
```

# Devtools

```{r devtools}
devtools::load_all()
```

## Command line arguments

```{r}
args <- list(optparse::make_option("--results", 
                                   type = "character", 
                                   default = "nested_xgboost_cox_cv_.*-20240220_095159.rds", 
                                   help = "Result iteration pattern [default %default]")) %>%
  optparse::OptionParser(option_list = .) %>% 
  optparse::parse_args()
```

```{r}
args %>% tibble::enframe(name = "Option", value = "Argument") %>% knitr::kable()
```

## Function factories

```{r}
mysave <- mysavefactory()
```

## Plot settings

```{r}
ggplot2::theme_set(ggthemes::theme_tufte(base_family = "sans", base_size = 12) +
                   ggplot2::theme(
                              panel.border = ggplot2::element_rect(colour = "black", fill = NA),
                              panel.background = ggplot2::element_rect(fill = "white", color = "white"),
                              plot.background = ggplot2::element_rect(fill = "white", color = "white"), 
                              axis.text = ggplot2::element_text(colour = "black", size = 10),
                              aspect.ratio = 1))
```

# Load previously run nested XGboost models

```{r}
result_files <- list.files(path = "rds", pattern = args$results, full.names = TRUE)
```

```{r}
result_files %>%
  tibble::as_tibble() %>% 
  DT::datatable()
```

```{r}
results <- result_files %>% 
  purrr::map(nested_xgboost_cox_cv_load) %>%
  rlang::set_names(., glue::glue("model_{seq_along(.)}")) 
```

## Harrell's statistic

```{r}
df_harrell <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}") %>% 
  purrr::map_df(nested_xgboost_cox_cv_harrell, .id = "model")
```

```{r}
df_harrell %>% DT::datatable()
```


```{r}
df_harrell %>%
  dplyr::summarize(mean = mean(index),
                   sd = sd(index)) %>%
  DT::datatable()
```

## Feature gain

```{r}
df_gain <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}") %>%
  purrr::map_df(~xgboost::xgb.importance(colnames(.x$matrix), .x$model), .id = "model")
```

```{r}
df_gain %>%
  dplyr::summarize(Gain = sum(Gain), .by = Feature) %>% 
  dplyr::arrange(desc(Gain)) %>%
  DT::datatable(caption = "Feature gain")
```

### Features better than randomness

```{r}
n_features <- df_gain %>%
  dplyr::summarize(Gain = sum(Gain), .by = Feature) %>% 
  dplyr::arrange(desc(Gain)) %>%
  dplyr::mutate(position = dplyr::row_number() - 1) %>% 
  dplyr::filter(stringr::str_detect(Feature, "rnorm_")) %>%
  dplyr::pull(position) %>%
  min()
```

Observed number of `r n_features` better than random features.

### Naming scheme

```{r}
df_importance <- df_gain %>%
  dplyr::mutate(model = stringr::str_replace(model, "_[0-9]*$", "")) %>% 
  dplyr::mutate(Gain_scaled = Gain/length(results))
```

```{r}
c_importance_order <- df_importance %>%
  dplyr::summarize(sum = sum(Gain_scaled), .by = Feature) %>%
  dplyr::arrange(sum) %>% 
  dplyr::top_n(n_features, wt = sum) %>% 
  dplyr::pull(Feature)
```

### XGBoost importance

```{r}
{ panel_a <- df_importance %>%
    dplyr::filter(Feature %in% c_importance_order) %>%
    dplyr::summarize(Gain_scaled = sum(Gain_scaled), .by = c(Feature, model)) %>%
    dplyr::mutate(Feature = factor(Feature, levels = c_importance_order)) %>% 
    ggplot2::ggplot(ggplot2::aes(x = Feature, y = Gain_scaled, fill = model, color = model)) +
    ggplot2::geom_bar(position='stack', stat = "identity") +
    ggplot2::coord_flip() +
    ggplot2::scale_color_grey(start = 0, end = 0.9) +  
    ggplot2::scale_fill_grey(start = 0, end = 0.9) +
    ggplot2::scale_x_discrete(name = NULL, expand = c(0, 0)) +
    ggplot2::scale_y_continuous(name = "Feature gain", expand = c(0, 0)) +
    ggplot2::theme(legend.position = "none",
                   panel.grid.major = ggplot2::element_blank(),
                   panel.grid.minor = ggplot2::element_blank()) } %>%
  mysave(name = "xgboost_importance", width = 12, height = 12)
```


## Calculate SHAPviz object

```{r}
calculate_shaps <- function(x, vars) {
  object <- shapviz::shapviz(x$model, X = x$data, X_pred = x$matrix)
  S <- shapviz::get_shap_values(object) %>%
    tibble::as_tibble() %>%
    tibble::add_column(ids = x$ids)
  X <- shapviz::get_feature_values(object) %>%
    tibble::as_tibble() %>%
    tibble::add_column(ids = x$ids)
  list(X = X, S = S)
}
```

```{r}
shap_interactions <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}") %>%
  purrr::map(calculate_shaps)
```

## SHAP beeswarm

```{r}
S_subset <- shap_interactions %>%
  purrr::map_df(~.x$S[, c_importance_order %union% "ids"], .id = "model") %>% 
  dplyr::summarize(dplyr::across(dplyr::one_of(c_importance_order), median), .by = ids) %>%
  tibble::column_to_rownames("ids") %>% 
  as.matrix()
```

```{r}
X_subset <- shap_interactions %>%
  purrr::map_df(~.x$X[, c_importance_order %union% "ids"], .id = "model") %>% 
  dplyr::summarize(dplyr::across(dplyr::one_of(c_importance_order), median), .by = ids) %>% 
  dplyr::mutate(dplyr::across(dplyr::one_of(c_importance_order), shapviz:::.min_max_scale)) %>%
  tibble::column_to_rownames("ids") %>% 
  as.matrix()
```

```{r}
df_shap <- as.data.frame.table(S_subset, responseName = "value") %>%
  dplyr::mutate(Feature = factor(Var2, levels = c_importance_order)) %>% 
  dplyr::mutate(color = as.data.frame.table(X_subset)$Freq) %>%
  tidyr::drop_na()
```

```{r}
panel_b <- ggplot2::ggplot(df_shap, ggplot2::aes(x = value, y = Feature)) +
  ggplot2::geom_vline(xintercept = 0, color = "darkgray") +
  ggplot2::geom_point(ggplot2::aes(color = color),
                      position = shapviz:::position_bee(width = 0.4, adjust = 0.5),
                      size = 1,
                      alpha = 1) +
  ggplot2::scale_colour_gradientn(colours = c("blue", "red"),
                                  breaks = c(0, 1),
                                  labels = c("low", "high")) +
  ggplot2::labs(x = "SHAP value", y = ggplot2::element_blank(), color = "Feature value")
```

```{r}
mysave(panel_b, name = "xgboost_beeswarm", width = 15, height = 10)
```


## Partial dependecy plot

```{r}
pdp_data_function <- function(x, results, ...) {
  feature <- x[[1]]
  model_number <- x[[2]]
  model <- results[[model_number]]$model
  vars <- results[[model_number]]$vars
  data <- results[[model_number]]$data %>%
    dplyr::select(one_of(vars))
  
  pdp::partial(model,
               pred.var = feature,
               train =  data,
               type = "regression",
               plot = FALSE) %>%
    dplyr::select(abundance = {{feature}}, yhat) %>%
    dplyr::mutate(model = model_number, Feature = feature)
}
```

```{r}
pdp_results <- results %>%
  purrr::list_flatten(name_spec = "{outer}_{inner}")
```

```{r}
c_factors <- X_subset %>%
  tibble::as_tibble() %>%
  dplyr::summarise_all(dplyr::n_distinct) %>%
  tidyr::gather(column, n_distinct) %>%
  dplyr::filter(n_distinct <= 10) %>%
  dplyr::pull(column)
```

```{r}
c_importance_order_cont <- c_importance_order %difference% c_factors 
```

```{r}
plot_data <- c_importance_order_cont %>% 
  purrr::cross2(seq_along(pdp_results)) %>%
  purrr::map_df(~pdp_data_function(.x, pdp_results))
```

```{r}
abund_min <- plot_data %>%
  dplyr::summarize(min = min(abundance), .by = c(Feature, model))

yhat_min <- dplyr::full_join(plot_data, abund_min, by = dplyr::join_by(Feature, model)) %>% 
  dplyr::filter(abundance == min) %>%
  dplyr::select(Feature, model, yhat_min = yhat)

plot_data_norm <- dplyr::full_join(plot_data, yhat_min, by = dplyr::join_by(Feature, model)) %>%
  dplyr::mutate(yhat_norm = (yhat - yhat_min)/yhat_min) %>%
  dplyr::mutate(abundance_norm = abundance) %>%
  dplyr::mutate(taxa = forcats::fct_rev(factor(Feature, levels = c_importance_order)))
```

```{r}
panel_c <- plot_data_norm %>%
  dplyr::filter(Feature %in% tail(c_importance_order_cont, 18)) %>% 
  ggplot2::ggplot(ggplot2::aes(x = abundance_norm, y = yhat_norm)) +
  ggplot2::facet_wrap(~forcats::fct_rev(Feature), scales = "free_x", ncol = 6) + 
  ggplot2::geom_line(ggplot2::aes(group = model), alpha = 0.2) +
  ggplot2::geom_hline(yintercept = 0) +
  ggplot2::geom_smooth(se=FALSE, colour="red", span=0.3, linetype="11", method = "gam") +
  ggplot2::scale_color_grey(start = 0, end = 0.9) +
  ggplot2::guides(color = "none") +
  ggplot2::scale_x_continuous(name = "Partial dependency", expand = c(0, 0)) +
  ggplot2::scale_y_continuous(name = NULL, expand = c(0, 0), lim = c(-1, 1)) +
  ggplot2::theme(strip.text.x = ggplot2::element_text(hjust = 0))
```

```{r}
mysave(panel_c, name = "xgboost_pdp", width = 15, height = 10)
```

# Panel plot

```{r}
layouts <- c(patchwork::area(t = 1, l = 1, b = 1, r = 1),
             patchwork::area(t = 1, l = 2, b = 1, r = 2),
             patchwork::area(t = 2, l = 1, b = 2 , r = 2))
```

```{r}
{ patchwork::wrap_plots(panel_a,
                        panel_b + ggplot2::theme(axis.text.y = ggplot2::element_blank()),
                        patchwork::free(panel_c)) +
    patchwork::plot_layout(design = layouts,
                           widths = c(1, 1),
                           heights = c(1, 1)) + 
    patchwork::plot_annotation(tag_levels = 'A') } %>%
  mysave(., name = "xgboost", width = 12, height = 10)
```
